Rapid Epidemiological Analysis of Comorbidities and Treatments as risk factors for COVID-19 in Scotland (REACT-SCOT): a population-based case-control study

### 2020-11-03 casecontrol.Rmd
multiple updates of pharmaco and shielding manuscripts

# rmarkdown::render("casecontrol.Rmd", output_file="casecontrol8June.pdf"); rmarkdown::render("response_reviewers.md")

@@

@@

_Methods and Findings_ -- The design was a matched case-control study. Severe COVID-19 was defined as either a positive nucleic acid test for SARS-CoV-2 in the national database followed by entry to a critical care unit or death within 28 days, or a death certificate with COVID-19 as underlying cause.  Up to ten controls per case matched for sex, age and primary care practice were selected from the population register.  All diagnostic codes from the past five years of hospitalisation records and all drug codes from prescriptions dispensed during the past nine months were extracted.  Rate ratios for severe COVID-19 were estimated by conditional logistic regression.

There were `r length(which(cc.severe$CASE==1))` severe cases. In a logistic regression using the age-sex distribution of the national population, the odds ratios for severe disease were `r round(exp(10 * logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.  In the case-control analysis, the strongest risk factor was residence in a care home, with rate ratio (95% CI) `r table.agegr.all[1, 3]`.  Univariate rate ratios (95% CIs) for conditions listed by public health agencies as conferring high risk were `r table.agegr.all["Type 1 diabetes", 3]` for Type 1 diabetes, `r table.agegr.all["Type 2 diabetes", 3]` for Type 2 diabetes, 
`r table.agegr.all["Ischaemic heart disease", 3]` for ischemic heart disease,
`r table.agegr.all["Other heart disease", 3]` for other heart disease, 
`r table.agegr.all["Asthma or chronic airway disease", 3]` for chronic lower respiratory tract disease, 
`r table.agegr.all["Chronic kidney disease or transplant recipient", 3]` for chronic kidney disease, 
`r table.agegr.all["Neurological (except epilepsy) or dementia", 3]` for neurological disease, 
`r table.agegr.all["Liver disease", 3]` for chronic liver disease and 
`r table.agegr.all["Immune deficiency or suppression", 3]` for immune deficiency or suppression.

`r gsub("([0-9]+ \\()(.*)(\\))", "\\2", freqs.all["listed.any", 2])` of cases and `r gsub("([0-9]+ \\()(.*)(\\))", "\\2", freqs.all["listed.any", 1])` of controls had at least one listed condition (`r gsub("([0-9]+ \\()(.*)(\\))", "\\2", table.agegr["Any listed condition", 2])` of cases and `r gsub("([0-9]+ \\()(.*)(\\))", "\\2", table.agegr["Any listed condition", 1])` of controls under age 40).  Severe disease was associated with encashment of at least one prescription in the past nine months and with at least one hospital admission in the past five years [rate ratios `r table.agegr.all["Any prescription", 3]`] and `r table.agegr.all["Any admission", 3]` respectively] even after adjusting for the listed conditions.  In those without listed conditions significant associations with severe disease were seen across many hospital diagnoses and drug categories. Age and sex provided `r round(cases.Lambda.agesex, 2)` bits of information for discrimination.  A model based on demographic variables, listed conditions, hospital diagnoses and prescriptions provided an additional `r table.densities[3, 5]` bits (C-statistic `r table.densities[3, 3]`).  A limitation of this study is that records from primary care were not available.

_Conclusions_ -- Along with older age and male sex, severe COVID-19 is strongly associated with past medical history across all age groups. Many comorbidities beyond the risk conditions designated by public health agencies contribute to this.  A risk classifier that uses all the information available in health records, rather than only a limited set of conditions, will more accurately discriminate between low-risk and high-risk individuals who may require shielding until the epidemic is over.

@@

# Author summary
Most people infected with the SARS-CoV-2 coronavirus do not become seriously ill.  It is The risk of severe or fatal illness is higher in older than in younger people, and is higher in people with conditions such as asthma and diabetes than in people without these conditions.  Using Scotland's capability for linking electronic health records, we report the first systematic study of the relation of severe or fatal COVID-19 to pre-existing health conditions and other risk factors.  We show that the strongest risk factor, apart from age, is residence in a care home. The conditions associated with increased risk include not only those already designated by public health agencies -- asthma, diabetes, heart disease, disabling neurological disease, kidney disease -- but many other diagnoses, associated with frailty and poor health.  This lays a basis for constructing risk scores based on electronic health records that can be used to advise people at high risk of severe disease to shield themselves when there cases in their neighbourhood.

# Background
Case series from many countries have suggested that in those with severe COVID-19 the prevalence of diabetes and cardiovascular disease is higher than expected.  For example in a large UK series the commonest co-morbidities were cardiac disease, diabetes, chronic pulmonary disease and asthma [@docherty_features_2020]. However there are also anecdotal reports of apparently healthy young persons succumbing to disease [@mcgoogan_why_2020].

Quantification of the risk associated with characteristics and co-morbidities has been limited by the lack of comparisons with the background population [@zhou_clinical_2020;@grasselli_baseline_2020;@guan_clinical_2020].  Two recent studies in the UK have included population comparators and have reported associations of in hospital test positive persons and COVID-19 death in hospital with co-morbidities including diabetes, asthma and heart disease [@niedzwiedz_ethnic_2020; @the_opensafely_collaborative_opensafely_2020].  These studies  have focused on conditions presumptively listed by public health agencies as increasing risk for COVID-19 based on case series data.

@@

# Methods
\hl{The study followed a pre-specified protocol. Modifications after study started were that we aligned the list of conditions of interest to be consistent with those designated as moderate risk conditions by public health agencies, and extended the list of drug classes under study to include all drugs.}

## Case definition and selection of matched controls
 All individuals testing positive for nucleic acid for SARS-CoV-2 were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database, which captures all virology testing in all NHS laboratories nationally. Linkage to other datasets was carried out using the Community Health Index (CHI) number, a unique identifier used in all care systems in Scotland.  Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units and has returned a daily census of patients in critical care from the beginning of the COVID-19 epidemic. Death registrations were obtained from linkage to the National Register of Scotland.  Severe or fatal COVID-19 was defined by either (1) a positive nucleic acid test followed by entry to critical care or death within 28 days; or (2) a death certificate with COVID-19 as underlying cause. \hl{Using this definition ensures ascertainment of all severe cases even if they die without testing positive or entering critical care, whatever selection policies may have limited entry to critical care.}

@@

## Morbidity and drug prescribing
For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-CoV-2 and using all codes on the discharge. Diagnostic coding under ICD chapters 5 (Mental, Behavioural and Neurodevelopmental) and 15 (Pregnancy) is incomplete as most psychiatric and maternity unit returns are not captured in SMR01. British National Formulary (BNF) drug codes for dispensed prescriptions issued in primary care were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016].  A cutoff date of 15 days before the incident date (date of testing positive for SARS-CoV-2, or 14 days before death for fatal cases without a positive test) was set, and prescriptions dispensed in a 240-day interval before this cutoff date were included. For this analysis prescription codes from BNF chapters 14 and above, comprising dressings, appliances, vaccines, anaesthesia and other preparations were grouped as "Other".

@@

Identifiable data were extracted by the Public Health Scotland CHI database and linkage team and de-identified before provision to the analysis team.

@@

## Risk factors
### Sociodemographic factors
Table \ref{tab:demog} shows univariate associations of demographic factors with severe disease.  Residence in a care home was by far the strongest risk factor for severe disease.  Higher risk of severe disease was also associated with socioeconomic deprivation.  In the   `r gsub("(.*\\()(.*)\\)", "\\2", table.ethnicsmr.missing[1, 2])` of cases  and  `r gsub("(.*\\()(.*)\\)", "\\2", table.ethnicsmr.missing[1, 1])` of controls  in whom ethnicity had been recorded in the Scottish Morbidity Record, there were few non-White cases or controls and the confidence limits for the risk ratios for severe or fatal disease by ethnic group were wide.

@@

## Comparison of fatal and non-fatal cases

Supplementary Table \ref{tab:casegr} shows a breakdown of severe cases by test-positive status, entry to critical care, and fatal versus non-fatal outcome.   \hl{Severe cases who entered critical care were much younger than severe cases never entering critical care.  The majority of severe cases from residential care homes never entered critical care. Among fatal cases who did not enter critical care. the distribution of age and other risk factors was similar in those with and without a positive test result, except that the proportion of care home residents was higher among those without a positive test result.   Among  those entering critical care  the median age and the proportion of males  and prevalence or prior comorbidities were higher in fatal than in non-fatal cases. }

## Systematic analysis of diagnoses associated with severe disease
The association of severe COVID-19 with prior hospital admission was examined further by testing for association of hospitalisations at each ICD-10 chapter level with severe COVID-19, among those without any of the listed conditions.  These results are shown in Supplementary Table \ref{tab:conditions}.  In univariate analyses, almost all ICD-10 chapters, with the exception of Chapters 7 (eye) Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.  In a multivariable analysis the strongest associations were with diagnoses in ICD chapters 4 (mental disorders) and 10 (respiratory). Supplementary Table \ref{tab:icdsubchapters} extracts univariate associations with ICD-10 subchapters in those without any listed conditions.  This table is filtered to show only subchapters for which there are at least 50 cases and controls and the univariate p-value is <0.001.  This shows that many subchapter diagnoses are associated with markedly higher risk of severe COVID-19.

@@

## Construction of a multivariable risk prediction model 
 The variables retained from the extended variable set (demographic variables, listed conditions, hospital diagnoses in each ICD-10 chapter, prescriptions in each BNF chapter) are shown in Table \ref{tab:stepwise}. Coefficients for specific conditions here should not be interpreted as effect estimates, as global variables for any hospital diagnosis and any listed condition have been included in the model. The predictive performance of the model chosen by stepwise regression was estimated by `r nfold`-fold cross-validation.  Observed and predicted case status were compared within each stratum over all test folds.  Table \ref{tab:xvalid} shows that using the extended set increased the C-statistic from `r table.densities[2, 3]` to `r table.densities[3, 3]` and the expected information for discrimination $\Lambda$ from `r table.densities[2, 5]` bits to `r table.densities[3, 5]` bits.

@@

# Discussion  
### Sociodemographic factors
This analysis confirms that risk for severe COVID-19 is associated with increasing age, male sex and socioeconomic deprivation.  The slope of the relationship of severe disease (on the scale of log odds) to age is less steep than the slope of the relationship of fatal disease to age.  <!---Another way of expressing this is that the fatality rate of severe disease increases with age. Although the relationships of mortality to age are linear and parallel for men and women, the fatality rate of severe disease is lower in women than in men up to age 50 and higher in women than in men between ages 50 and 70.  This may reflect either variation in the biology of the disease, or differential selection for entry to critical care.--->  Residence in a care home was associated with a `r round(as.numeric(substr(table.agegr.all[1, 3], 1, 4)))`-fold increased rate of severe COVID-19 in this age matched analysis, reduced to `r round(as.numeric(substr(table.agegr.all[1, 5], 1, 4)))`-fold by adjustment for listed conditions. This excess risk is likely to reflect both the spread of the epidemic in care homes and residual confounding by frailty.

Although the assignment of ethnicity is incomplete and the prevalence of non-White ethnic groups in Scotland is low, we show the data as they give some indication of the likely upper bound of the absolute numbers of severe cases in South Asian and Black ethnic groups in Scotland. This is relevant because substantial elevations in risk of hospitalised and fatal COVID-19 have been reported elsewhere in the UK [@the_opensafely_collaborative_opensafely_2020; @niedzwiedz_ethnic_2020; @ho_modifiable_2020]. In the OpenSAFELY study risk ratios for fatal COVID-19 of 1.7 in those recorded as Black and and 1.6 in those recorded as Asian, in comparison with those recorded as White, persisted after adjustment for comorbidities and socioeconomic status. In Scotland however, although point estimates for rate ratios in those of South Asian and Black ethnicity compared to Whites are above 1, the confidence intervals are too wide for any definitive conclusions on ethnic differences.

### Co-morbidities
We have confirmed that the moderate risk conditions designated by the NHS and other agencies [@nhs_whos_2020] are associated with increased risk of severe COVID-19.  However the rate ratios associated with these conditions vary with age - for example the rate ratio associated with diabetes is higher at younger ages. The rate ratios of `r round(as.numeric(substr(table.agegr.all["Type 1 diabetes", 3], 1, 4)), 1)` for Type 1 diabetes and `r round(as.numeric(substr(table.agegr.all["Type 2 diabetes", 3], 1, 4)), 1)` for Type 2 diabetes are broadly similar to those reported in UK Biobank and in the OpenSAFELY studies.  We confirm the higher risk with asthma and chronic lung disease and liver disease reported in these and earlier studies.  An unexpected finding was that the risk associated with other forms of heart disease is higher than that associated with ischaemic heart disease.  This category includes conditions such as atrial fibrillation, cardiomyopathies and heart failure. One of the highest rate ratios is that associated with chronic kidney disease.  Prevention of nosocomial transmission in dialysis units may help to reduce this risk.  Over all age groups, `r gsub("(.*\\()(.*)\\)", "\\2", freqs.all["listed.any", 2])` of severe cases had at least one of these listed conditions. \hl{ In this dataset it is not possible to examine adequately the risk associated with neoplasms, as we cannot separately identify those who were advised to shield themselves because they had active neoplasms of lymphoid or hematopoietic tissue or were receiving treatments that affect the immune system. We hope to explore this in a separate study that will include linkage to records of shielding advice.} In patients without any listed conditions, further systematic evaluation of past hospitalisation history did not reveal a sparse set of underlying conditions; instead many diagnoses were associated with severe COVID-19.

Media reports of apparently healthy young people succumbing to severe COVID-19 have disseminated the message that all are at risk of disease whatever their age or health status.  However we found that half of cases under 40 years had at least one of the listed conditions and among those who did not have one of these conditions, the proportions who had at least one prior hospitalisation or dispensed prescription were higher in cases than in controls.  In all age groups, very few of the fatal cases had not had either a hospital admission in the past five years or a dispensed prescription in the past year.

A striking finding of this study was the association of severe COVID-19 with dispensing of at least one prescription in the past year, only partly explained by higher rates of prescribing among those with listed conditions.  Partitioning of this association between BNF chapters, which represent broad indication-based drug classes, showed that prescription of drugs for the gastrointestinal and central nervous systems, together with nutritional supplements, contributed to this association.  Although it is likely that most associations of severe COVID-19 with drug prescribing are attributable to the indications for which these drugs were prescribed, or to more diffuse frailty especially in older persons, causal effects of drugs or direct effects of polypharmacy on susceptibility cannot be ruled out.  These associations are explored in a separate paper.

## Relevance to policy
As lockdown restrictions are eased, there is general agreement that vulnerable individuals will require shielding, even if the restart of the epidemic can be slowed or suppressed by mass testing, contact tracing and isolation of those who test positive.  The "stratify and shield" policy option [@mckeigue_evaluation_2020], in which high-risk individuals comprising up to 15% of the population are shielded for a defined period while the epidemic is allowed to run relatively quickly in low-risk individuals until population-level immunity is attained, depends critically on informative risk discrimination.  So too does the similarly named "segment and shield " option [@bunnik_segmentation_2020] which has the opposite objective of keeping transmissions low.  Although for this preliminary study we have not used the full repertoire of machine learning methods available for this type of problem, we have shown that a model based on health records provides `r table.densities[3, 5]` bits of information for discrimination conditional on age and sex.  Adding this to the `r round(cases.Lambda.agesex, 2)` bits provided by age and sex gives a total information for discrimination of `r round(table.densities[3, 5] + cases.Lambda.agesex, 1)` bits.  \hl{We have shown elsewhere that this level of predictive performance would allow at least 80% of those at risk of severe or fatal disease to be allocated to a shielded group comprising no more than 15% of the population} [@mckeigue_evaluation_2020]

>>>>

# rmarkdown::render("casecontrol.Rmd", output_file="casecontrol_supptables.pdf"); rmarkdown::render("response_reviewers.md")

@@

pctonly <- function(x) gsub("([0-9]+ \\()(.*)(\\))", "\\2", x)

rr.pval <- function(r, p) {
    r <- gsub(",", " to", r)
    p <- gsub("ensuremath\\{", "\\ensuremath\\{p=", p)
    p <- gsub("times", "\\\\times", p)
    gsub("\\)", paste0(", \\\\", p, "\\)"), r)
}

##rr.pval(table.agegr.all[1, 3], table.agegr.all[1, 4])

@@

_Methods and Findings_ -- The design was a matched case-control study. Severe COVID-19 was defined as either a positive nucleic acid test for SARS-CoV-2 in the national database followed by entry to a critical care unit or death within 28 days, or a death certificate with COVID-19 as underlying cause.  Up to ten controls per case matched for sex, age and primary care practice were selected from the national population register.  For this analysis based on ascertainment of positive test results up to `r gsub("^0", "", format(max(cc.severe$SPECIMENDATE, na.rm=TRUE), "%d %B %Y"))`, entry to critical care up to `r gsub("^0", "", format(maxdate.sicsag, "%d %B %Y"))` and deaths registered up to `r format(max(cc.severe$DATE_OF_REGISTRATION, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[CASE==0])` controls and `r nrow(cc.severe[CASE==1])` cases of whom `r table.demog.append["Care home", 3]` were care home residents.  All diagnostic codes from the past five years of hospitalisation records and all drug codes from prescriptions dispensed during the past 240 days were extracted.  Rate ratios for severe COVID-19 were estimated by conditional logistic regression.

In a logistic regression using the age-sex distribution of the national population, the odds ratios for severe disease were `r round(exp(10 * logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.  In the case-control analysis, the strongest risk factor was residence in a care home, with rate ratio (95% CI) `r rr.pval(table.agegr.all[1, 3], table.agegr.all[1, 4])`.  Univariate rate ratios (95% CIs) for conditions listed by public health agencies as conferring high risk were `r rr.pval(table.agegr.all["Type 1 diabetes", 3], table.agegr.all["Type 1 diabetes", 4])` for Type 1 diabetes, `r rr.pval(table.agegr.all["Type 2 diabetes", 3], table.agegr.all["Type 2 diabetes", 4])` for Type 2 diabetes, 
`r rr.pval(table.agegr.all["Ischaemic heart disease", 3], table.agegr.all["Ischaemic heart disease", 4])` for ischemic heart disease,
`r rr.pval(table.agegr.all["Other heart disease", 3], table.agegr.all["Other heart disease", 4])` for other heart disease, 
`r rr.pval(table.agegr.all["Asthma or chronic airway disease", 3], table.agegr.all["Asthma or chronic airway disease", 3])` for chronic lower respiratory tract disease, 
`r rr.pval(table.agegr.all["Chronic kidney disease or transplant recipient", 3], table.agegr.all["Chronic kidney disease or transplant recipient", 4])` for chronic kidney disease, 
`r rr.pval(table.agegr.all["Neurological (except epilepsy) or dementia", 3], table.agegr.all["Neurological (except epilepsy) or dementia", 4])` for neurological disease, 
`r rr.pval(table.agegr.all["Liver disease", 3], table.agegr.all["Liver disease", 4])` for chronic liver disease and 
`r rr.pval(table.agegr.all["Immune deficiency or suppression", 3], table.agegr.all["Immune deficiency or suppression", 4])` for immune deficiency or suppression.

`r pctonly(freqs.all["listed.any", 2])` of cases and `r pctonly(freqs.all["listed.any", 1])` of controls had at least one listed condition (`r pctonly(table.agegr["Any listed condition", 2])` of cases and `r pctonly(table.agegr["Any listed condition", 1])` of controls under age 40).  Severe disease was associated with encashment of at least one prescription in the past nine months and with at least one hospital admission in the past five years [rate ratios `r table.agegr.all["Any prescription", 3]`] and `r table.agegr.all["Any admission", 3]` respectively] even after adjusting for the listed conditions.  In those without listed conditions significant associations with severe disease were seen across many hospital diagnoses and drug categories. Age and sex provided `r round(cases.Lambda.agesex, 2)` bits of information for discrimination.  A model based on demographic variables, listed conditions, hospital diagnoses and prescriptions provided an additional `r table.densities[3, 5]` bits (C-statistic `r table.densities[3, 3]`).  A limitation of this study is that records from primary care were not available.

_Conclusions_ -- We have shown that along with older age and male sex, severe COVID-19 is strongly associated with past medical history across all age groups. Many comorbidities beyond the risk conditions designated by public health agencies contribute to this.  A risk classifier that uses all the information available in health records, rather than only a limited set of conditions, will more accurately discriminate between low-risk and high-risk individuals who may require shielding until the epidemic is over.

@@

# Author summary
## Why was this study done? 
* Most people infected with the SARS-CoV-2 coronavirus do not become seriously ill: risk of severe or fatal disease is associated with older age, male sex, and conditions designated by public health agencies including asthma, diabetes and heart disease.

* Studies reported so far have focused on these "listed conditions", but have not examined medical records systematically to identify possible risk factors for severe COVID-19.

* The objectives of this study were to identify risk factors for severe COVID-19 and to lay the basis for risk stratification based on electronic health records.

## What did the researchers do and find? 
* Using Scotland's capability for linking electronic health records, we report the first systematic study of the relation of severe or fatal COVID-19 to pre-existing health conditions and other risk factors.

* Residents in care homes were 21 times more likely to develop severe disease than people of the same age and sex not living in care homes.

* The conditions associated with increased risk include not only those already designated by public health agencies -- asthma, diabetes, heart disease, disabling neurological disease, kidney disease -- but other diagnoses that are associated with frailty and poor health such as strokes and a history of falls.

* In those without any listed conditions, use of prescribed drugs acting on the digestive system or nervous system is associated with increased risk of severe COVID-19.

## What do these findings mean? 
* The risk to younger individuals without any recent history of hospital admission or use of prescription drugs is very low.

* This study lays a basis for calculating a risk score based on electronic health records for every individual in the population, and using it to advise those at high risk of severe disease to shield themselves when there is a COVID-19 epidemic in their locality.

# Background
Case series from many countries have suggested that in those with severe COVID-19 the prevalence of diabetes and cardiovascular disease is higher than expected.  For example in a large UK series the commonest co-morbidities were cardiac disease, diabetes, chronic pulmonary disease and asthma [@docherty_features_2020-1]. However there are also anecdotal reports of apparently healthy young persons succumbing to disease [@mcgoogan_why_2020].

Quantification of the risk associated with characteristics and co-morbidities has been limited by the lack of comparisons with the background population [@zhou_clinical_2020;@grasselli_baseline_2020;@guan_clinical_2020].  Two recent studies in the UK have included population comparators and have reported associations of in hospital test positive persons and COVID-19 death in hospital with co-morbidities including diabetes, asthma and heart disease [@niedzwiedz_ethnic_2020; @williamson_opensafely_2020].  These studies  have focused on conditions presumptively listed by public health agencies as increasing risk for COVID-19 based on case series data.

@@

# Methods
The protocol of the study dated 16 April 2020, together with all code used to prepare this manuscript is available in a public [repository](https://github.com/pmckeigue/covid-scotland_public/).  We modified the original protocol to align the list of risk conditions to be consistent with those designated by public health agencies, and extended the list of drug classes under study to include all drugs. The study was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number [EUPAS35558](http://www.encepp.eu/encepp/viewResource.htm?id=35559)). All record linkage studies using NHS data in Scotland are governed by the Public Benefit and Privacy Panel for Health and Social Care, which includes patient and public representatives.  Identifiable data were extracted by the Public Health Scotland CHI database and linkage team and de-identified before provision to the analysis team.

## Case definition and selection of matched controls
 All individuals testing positive for nucleic acid for SARS-CoV-2 were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database, which captures all virology testing in all NHS laboratories nationally. Linkage to other datasets was carried out using the Community Health Index (CHI) number, a unique identifier used in all care systems in Scotland.  Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units and has returned a daily census of patients in critical care from the beginning of the COVID-19 epidemic. Death registrations were obtained from linkage to the National Register of Scotland.  Severe or fatal COVID-19 was defined by either (1) a positive nucleic acid test followed by entry to critical care or death within 28 days; or (2) a death certificate with COVID-19 as underlying cause. Using this definition ensures ascertainment of all severe cases even if they die without testing positive or entering critical care, whatever selection policies may have limited entry to critical care.

@@

## Morbidity and drug prescribing
For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-CoV-2 and using all codes on the discharge. Diagnostic coding under ICD chapters V (Mental, Behavioural and Neurodevelopmental) and XV (Pregnancy) is incomplete as most psychiatric and maternity unit returns are not captured in SMR01. British National Formulary (BNF) drug codes for dispensed prescriptions issued in primary care were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016].  A cutoff date of 15 days before the incident date (date of testing positive for SARS-CoV-2, or 14 days before death for fatal cases without a positive test) was set, and prescriptions dispensed in a 240-day interval before this cutoff date were included. For this analysis prescription codes from BNF chapters 14 and above, comprising dressings, appliances, vaccines, anaesthesia and other preparations were grouped as "Other".

@@

@@

## Risk factors
### Sociodemographic factors
Table \ref{tab:demog} shows univariate associations of demographic factors with severe disease.  Residence in a care home was by far the strongest risk factor for severe disease.  Higher risk of severe disease was also associated with socioeconomic deprivation.  In the   `r gsub("(.*\\()(.*)\\)", "\\2", table.ethnicsmr.missing[1, 2])` of cases  and  `r gsub("(.*\\()(.*)\\)", "\\2", table.ethnicsmr.missing[1, 1])` of controls  in whom ethnicity had been recorded in the Scottish Morbidity Record, there were few non-White individuals and the confidence limits for the rate ratios by ethnic group were wide.

@@

## Comparison of fatal and non-fatal cases
Supplementary Table \ref{tab:casegr} shows a breakdown of severe cases by test-positive status, entry to critical care, and fatal versus non-fatal outcome.  Severe cases who entered critical care were much younger than severe cases never entering critical care.  The majority of severe cases from residential care homes never entered critical care. Among fatal cases who did not enter critical care. the distribution of age and other risk factors was similar in those with and without a positive test result, except that the proportion of care home residents was higher among those without a positive test result.  Among those entering critical care the median age and the proportion of males and prevalence or prior comorbidities were higher in fatal than in non-fatal cases.

## Systematic analysis of diagnoses associated with severe disease
The association of severe COVID-19 with prior hospital admission was examined further by testing for association of hospitalisations at each ICD-10 chapter level with severe COVID-19, among those without any of the listed conditions.  These results are shown in Supplementary Table \ref{tab:conditions}.  In univariate analyses, almost all ICD-10 chapters, with the exception of Chapters VII (eye), VIII (ear) and XV (pregnancy) were associated with increased risk of severe disease.  In a multivariable analysis the strongest associations were with diagnoses in ICD chapters IV (mental disorders) and X (respiratory). Supplementary Table \ref{tab:icdsubchapters} extracts univariate associations with ICD-10 subchapters in those without any listed conditions.  This table is filtered to show only subchapters for which there are at least 50 cases and controls and the univariate p-value is <0.001.  This shows that many subchapter diagnoses are associated with markedly higher risk of severe COVID-19.

@@

## Construction of a multivariable risk prediction model 
 The variables retained from the extended variable set (demographic variables, listed conditions, hospital diagnoses in each ICD-10 chapter, prescriptions in each BNF chapter) are shown in Table \ref{tab:stepwise}. Coefficients for specific conditions here should not be interpreted as effect estimates, as global variables for any hospital diagnosis and any listed condition have been included in the model. The predictive performance of the model chosen by stepwise regression was estimated by `r nfold`-fold cross-validation.  Observed and predicted case status were compared within each stratum over all test folds.  Table \ref{tab:xvalid} shows that in comparison with using only demographic variables and listed conditions, using the extended variable set increased the C-statistic from `r table.densities[2, 3]` to `r table.densities[3, 3]` and the expected information for discrimination $\Lambda$ from `r table.densities[2, 5]` bits to `r table.densities[3, 5]` bits.

@@

# Discussion  
## Sociodemographic factors
This analysis confirms that risk for severe COVID-19 is associated with increasing age, male sex and socioeconomic deprivation.  The slope of the relationship of severe disease (on the scale of log odds) to age is less steep than the slope of the relationship of fatal disease to age.  <!---Another way of expressing this is that the fatality rate of severe disease increases with age. Although the relationships of mortality to age are linear and parallel for men and women, the fatality rate of severe disease is lower in women than in men up to age 50 and higher in women than in men between ages 50 and 70.  This may reflect either variation in the biology of the disease, or differential selection for entry to critical care.--->  Residence in a care home was associated with a `r round(as.numeric(substr(table.agegr.all[1, 3], 1, 4)))`-fold increased rate of severe COVID-19 in this age matched analysis, reduced to `r round(as.numeric(substr(table.agegr.all[1, 5], 1, 4)))`-fold by adjustment for listed conditions. This excess risk is likely to reflect both the spread of the epidemic in care homes and residual confounding by frailty.

As the proportion of the Scottish population that is of non-White ethnicity is low and the assignment of ethnicity in this dataset is incomplete, the confidence intervals for the rate ratios associated with South Asian and Black ethnicity are wide.  Studies from England [@williamson_opensafely_2020; @niedzwiedz_ethnic_2020; @ho_modifiable_2020] have reported elevations in risk of hospitalised and fatal COVID-19 in non-White ethnic groups; in the OpenSAFELY study the risk ratios for fatal COVID-19 associated with Black and Asian ethnicity were 1.7 and 1.6 respectively.  The confidence intervals in this study are compatible with the effect sizes estimated in England.

## Co-morbidities
We have confirmed that the moderate risk conditions designated by the NHS and other agencies [@nhs_whos_2020] are associated with increased risk of severe COVID-19.  The rate ratios of `r round(as.numeric(substr(table.agegr.all["Type 1 diabetes", 3], 1, 4)), 1)` for Type 1 diabetes and `r round(as.numeric(substr(table.agegr.all["Type 2 diabetes", 3], 1, 4)), 1)` for Type 2 diabetes are broadly similar to those reported in the UK Biobank [@ho_modifiable_2020] and OpenSAFELY [@williamson_opensafely_2020] studies.  We confirm the higher risk with asthma and chronic lung disease and liver disease reported in these and earlier studies.  The rate ratios associated with these risk conditions vary with age: for example the rate ratio associated with diabetes is higher at younger ages. An unexpected finding was that the risk associated with other forms of heart disease is higher than that associated with ischaemic heart disease.  This category includes conditions such as atrial fibrillation, cardiomyopathies and heart failure. One of the highest rate ratios is that associated with chronic kidney disease.  Prevention of nosocomial transmission in dialysis units may help to reduce this risk.  Over all age groups, `r gsub("(.*\\()(.*)\\)", "\\2", freqs.all["listed.any", 2])` of severe cases had at least one of these listed conditions. In this dataset it is not possible to examine adequately the risk associated with neoplasms as we cannot separately identify those who were advised to shield themselves because they had active neoplasms of lymphoid or hematopoietic tissue or were receiving treatments that affect the immune system. We plan to explore this in a separate study based on linkage to records of shielding advice.  In patients without any listed conditions, further systematic evaluation of past hospitalisation history did not reveal a sparse set of underlying conditions; instead many diagnoses were associated with severe COVID-19.

Public health agencies [@ghebreyesus_who_2020] and media reports of apparently healthy young people succumbing to severe COVID-19 [@mcgoogan_why_2020] have disseminated the message that all are at risk of severe COVID-19 whatever their age or health status.  However we found that half of cases under 40 years had at least one of the listed conditions and among those who did not have one of these conditions, the proportions who had at least one prior hospitalisation or dispensed prescription were higher in cases than in controls.  In all age groups, very few of the fatal cases had not had either a hospital admission in the past five years or a dispensed prescription in the past year.

A striking finding of this study was the association of severe COVID-19 with dispensing of at least one prescription in the 240-day interval preceding the cutoff of 15 days before diagnosis, only partly explained by higher rates of prescribing among those with listed conditions.  Partitioning of this association between BNF chapters, which represent broad indication-based drug classes, showed that prescription of drugs for the gastrointestinal and central nervous systems, together with nutritional supplements, contributed to this association.  Although it is likely that most associations of severe COVID-19 with drug prescribing are attributable to the indications for which these drugs were prescribed, or to more diffuse frailty especially in older persons, causal effects of drugs or direct effects of polypharmacy on susceptibility cannot be ruled out.  These associations are explored in a separate paper.

## Relevance to policy
As lockdown restrictions are eased, there is general agreement that vulnerable individuals will require shielding, even if the restart of the epidemic can be slowed or suppressed by mass testing, contact tracing and isolation of those who test positive.  The "stratify and shield" policy option [@mckeigue_evaluation_2020], in which high-risk individuals comprising up to 15% of the population are shielded for a defined period while the epidemic is allowed to run relatively quickly in low-risk individuals until population-level immunity is attained, depends critically on informative risk discrimination.  So too does the similarly named "segment and shield " option [@bunnik_segmentation_2020] which has the opposite objective of keeping transmissions low.  Although in this preliminary study we have not used the full repertoire of machine learning methods available for constructing predictive models, we have shown that a model based on health records provides `r table.densities[3, 5]` bits of information for discrimination conditional on age and sex.  Adding this to the `r round(cases.Lambda.agesex, 2)` bits provided by age and sex gives a total information for discrimination of `r round(table.densities[3, 5] + cases.Lambda.agesex, 1)` bits.  We have shown elsewhere that this level of predictive performance would allow at least 80% of those at risk of severe or fatal disease to be allocated to a shielded group comprising no more than 15% of the population [@mckeigue_evaluation_2020].

As awareness grows of how risk varies between individuals, individuals will seek information about their own level of risk.  A key implication of our results is that risk of severe or fatal disease is multifactorial.  The rate ratio of `r round(exp(10 * logistic.coeffs[3, 1]), 1)` associated with a 10-year increase in age is stronger than the rate ratios associated with common diseases such as asthma or Type 2 diabetes that are listed as conditions associated with high risk.  A corollary of this is that a crude classification based on assigning all persons with a listed condition to a group for whom shielding is recommended will have poor specificity, as one quarter of those aged 60-74 years in the population have at least one of the listed conditions we examined.  It will also exclude many people at high risk because they have multiple risk factors each of small effect. A more meaningful way to score risk for an individual would be to use all available information to calculate a "COVID age" as the age at which the average risk for someone of the same sex in the population equates to the risk for the individual under study. Thus the rate ratio of `r rr <- round(as.numeric(gsub(" .+", "", table.agegr.all["Type 1 diabetes", 3])), 1); rr` associated with Type 1 diabetes equates to an increase of `r round(log(rr) / logistic.coeffs[3, 1], 1)` years in COVID age.  In Scotland it is technically possible to use existing electronic health records to calculate a risk score for every individual in the population, though more work would be required to develop this as a basis for official advice and individual decisions.
